\section{Introduction}
\textsc{Phi} is a constructed language. A constructed language is one whose
phonology, grammar, and vocabulary, instead of having developed naturally, are
consciously devised.\footnote{\url{https://en.wikipedia.org/wiki/Constructed%
\_language}} The sound of the name \emph{Phi} embodies the simplicity and
softness of the language itself. \emph{Phi~($\varPhi$)} is also beauty in
mathematics. It is

\begin{quote}
  \emph{the golden ratio, an irrational number with a value of approximately
  1.618033988 which expresses the relationship that the sum of two quantities
  is to the larger quantity as the larger is to the smaller.} -- Wiktionary%
  \footnote{\url{https://en.wiktionary.org/wiki/\%CF\%86}}
\end{quote}

It is thought that the value of \emph{phi} is one of the fundamental
characteristics of nature; in the language Phi, nature is fundamental to its
being. The natural world is the root of many of the words in the Phi lexicon.
More industrial or unnatural words are constructed by describing the subject in
terms of existing, more natural words. As an historical (or perhaps Hollywood)
example, native North Americans called a rifle a \emph{fire stick}. They
described something in terms with which they were familiar.

There is also an emphasis on simplicity in the language. Phi has many aspects of
a pidgin language without having been derived from existing languages. An a
priori pidgin is definitely a contradiction in terms, but the resemblance to a
pidgin is intentional. Wikipedia describes some of the common traits of such a
language.%
\footnote{\url{https://en.wikipedia.org/wiki/Pidgin\#Common_traits}}
These are the similarities in Phi to those traits:

\begin{itemize}[label=\raisebox{0.33ex}{\tiny$\bullet$},topsep=0pt]\itemsep-5pt
  \item Phi is a isolating language, but not an extreme example of one;
  \item It is also an analytic language, but taken further than most pidgins;
  \item There are no embedded clauses in Phi;
  \item Phi eschews syllable codas;
  \item Consonant clusters are restricted to voiceless digraphs,
  e.g. \dscopy /sh/\normalfont;
  \item Aspiration only occurs with initial \dscopy /p/~\normalfont and
  \dscopy /t/\normalfont;
  \item Phi uses the common five-vowel system: \dscopy /i/, /u/, /e/, /o/, /a/%
  \normalfont;
  \item There is no sound influence between morphemes;
  \item There are no tones in Phi;
  \item Phi has no grammatical tense;
  \item No conjugation or declension occurs;
  \item There is a lack of grammatical gender or number.
\end{itemize}

These traits are where Phi diverges from a pidgin:

\begin{itemize}[label=\raisebox{0.33ex}{\tiny$\bullet$},topsep=0pt]\itemsep-5pt
  \item There are no historical sound changes in an a priori language;
  \item No monopthongization occurs in Phi: each vowel is pronounced separately;
  \item Phi has clear parts of speech and word categorization;
  \item In addition, Phi uses a SOV word order.
\end{itemize}

Phi is an attempt to answer a few questions. What if humans evolved to cooperate
with nature instead of trying to dominate it? What if language remained simple
and straightforward instead of growing in complexity as society advanced? What
if society naturally tended toward the positive instead of the negative? What
would that language look and sound like? In this respect Phi can be seen as an
engineered language, specifically a philosophical language.%
\footnote{\url{https://en.wikipedia.org/wiki/Philosophical_language}}